---
title: "Clustering Lab"
author: "Courtney Kennedy"
date: "9/20/2020"
output: html_document
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r echo=FALSE}
library(e1071)
library(tidyverse)
library(plotly)
library(htmltools)
library(devtools)
library(caret)
library(NbClust)

```
```{r}
house_votes_Rep = read_csv("~/Fall21/introDS/DS-3001-New/data/house_votes_Rep.csv")

table(house_votes_Rep$party.labels)
#View(house_votes_Rep)
```

Goal: Know how to make decisions and answer questions using clustering. 

Repeat the clustering process only using the Rep house votes dataset
- What differences and similarities did you see between how the clustering 
worked for the datasets?

The main difference is now the republican cluster is the one with more aye votes and less nay votes vs the dem house votes dataset has the democratic cluster with more aye votes and less nay votes.  

```{r}
#Select the variables to be included in the cluster 
clust_data_Rep = house_votes_Rep[, c("aye", "nay", "other")]
```

```{r}
#Run the clustering algo with 2 centers
set.seed(1)
kmeans_obj_Rep = kmeans(clust_data_Rep, centers = 2, 
                        algorithm = "Lloyd")
```

```{r}
#View the results
kmeans_obj_Rep
```

```{r}
#Visualize the output
party_clusters_Rep = as.factor(kmeans_obj_Rep$cluster)
ggplot(house_votes_Rep, aes(x = aye, 
                            y = nay,
                            color = party.labels,  #<- tell R how to color 
                            #   the data points
                            shape = party_clusters_Rep)) + 
  geom_point(size = 6) +
  ggtitle("Aye vs. Nay votes for Republican-introduced bills") +
  xlab("Number of Aye Votes") +
  ylab("Number of Nay Votes") +
  scale_shape_manual(name = "Cluster", 
                     labels = c("Cluster 1", "Cluster 2"),
                     values = c("1", "2")) +
  scale_color_manual(name = "Party",         #<- tell R which colors to use and
                     #   which labels to include in the legend
                     labels = c("Deomcratic", "Republican"),
                     values = c("blue", "red")) +
  theme_light()


```

```{r}
#save as a png
ggsave("US House Votes for Rep Bills.png", 
       width = 10, 
       height = 5.62, 
       units = "in")
```

```{r}
#Evaluate the quality of the clustering 

# Inter-cluster variance,
# "betweenss" is the sum of the distances between points 
# from different clusters.
num_Rep = kmeans_obj_Rep$betweenss

# Total variance, "totss" is the sum of the distances
# between all the points in the data set.
denom_Rep = kmeans_obj_Rep$totss

# Variance accounted for by clusters.
(var_exp_Rep = num_Rep / denom_Rep)
```

```{r}
#Use the function we created to evaluate several different number of clusters

# The function explained_variance wraps our code for calculating 
# the variance explained by clustering.
explained_variance = function(data_in, k){
  
  # Running the kmeans algorithm.
  set.seed(1)
  kmeans_obj = kmeans(data_in, centers = k, algorithm = "Lloyd", iter.max = 30)
  
  # Variance accounted for by clusters:
  # var_exp = intercluster variance / total variance
  var_exp = kmeans_obj$betweenss / kmeans_obj$totss
  var_exp  
}
explained_var_Rep = sapply(1:10, explained_variance, data_in = clust_data_Rep)
#View(explained_var_Rep)
```

```{r}
#Create a elbow chart of the output 

# Data for ggplot2.
elbow_data_Rep = data.frame(k = 1:10, explained_var_Rep)
#View(elbow_data_Rep)

# Plotting data.
ggplot(elbow_data_Rep, 
       aes(x = k,  
           y = explained_var_Rep)) + 
  geom_point(size = 4) +           #<- sets the size of the data points
  geom_line(size = 1) +            #<- sets the thickness of the line
  xlab('k') + 
  ylab('Inter-cluster Variance / Total Variance') + 
  theme_light()
```

```{r}
#Use NbClust to select a number of clusters
# Run NbClust.
(nbclust_obj_Rep = NbClust(data = clust_data_Rep, method = "kmeans"))

# View the output of NbClust.
nbclust_obj_Rep

# View the output that shows the number of clusters each method recommends.
#View(nbclust_obj_Rep$Best.nc)
```

```{r}
#Display the results visually 
freq_k_Rep = nbclust_obj_Rep$Best.nc[1,]
freq_k_Rep = data.frame(freq_k_Rep)
#View(freq_k_Rep)

# Check the maximum number of clusters suggested.
max(freq_k_Rep)

#essentially resets the plot viewer back to default
#dev.off()

# Plot as a histogram.
ggplot(freq_k_Rep,
       aes(x = freq_k_Rep)) +
  geom_bar() +
  scale_x_continuous(breaks = seq(0, 15, by = 1)) +
  scale_y_continuous(breaks = seq(0, 12, by = 1)) +
  labs(x = "Number of Clusters",
       y = "Number of Votes",
       title = "Cluster Analysis")
```

```{r}
#Using the recommended number of cluster compare the quality of the model 
#with 2 clusters 
# Both the elbow graph and the nbc Cluster method recommend two clusters.  
```


```{r}
#Bonus: Create a 3d version of the output

party_color3D_Rep = data.frame(party.labels = c("Democrat", "Republican"),
                               color = c("blue", "red"))

View(party_color3D_Rep)


# Join the new data frame to our house_votes_Dem data set.
house_votes_color_Rep = inner_join(house_votes_Rep, party_color3D_Rep)

house_votes_color_Rep$clusters <- (party_clusters_Rep)

str(house_votes_color_Rep)

#Remove special characters
house_votes_color_Rep$Last.Name <- gsub("[^[:alnum:]]", "", house_votes_color_Rep$Last.Name)

# Use plotly to do a 3d imaging 

fig <- plot_ly(house_votes_color_Rep, 
               type = "scatter3d",
               mode="markers",
               symbol = ~clusters,
               x = ~aye, 
               y = ~nay, 
               z = ~other,
               color = ~color,
               colors = c('#0C4B8E','#BF382A'), 
               text = ~paste('Representative:',Last.Name,
                             "Party:",party.labels))


fig
# dev.off()

#Flat so the other category isn't impacting the clusters very much 
```


  





